---
title: "ViewPoint.ca Listings"
# author: "Colin Douglas"
date: Last Updated `r format(Sys.time(), "%b %d, %Y at %H:%M %p")`
always_allow_html: true
permalink: /real-estate/
layout: page
excerpt: Aggregated data on Nova Scotia real estate listings
comments: false
output: 
  md_document:
    variant: markdown_phpextra
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.path="figures/")

library(tidyverse)
library(lubridate)
library(kableExtra)
library(broom)
source("setup.R")
```

```{r Read in data, define regions}

listings_all <- read_csv("data/listings-clean.csv", col_types = cols())

listings <- listings_all %>%
  filter(!is.na(type), !is.na(status)) %>%
  arrange(datetime) %>%
  # Keep only the unique rows, because sometimes things get posted more than once
  distinct(mls_no, pid, price, status, .keep_all = TRUE) %>%
  mutate(status = factor(status, levels = names(status_colors)),
         type = factor(case_when(
           type == "Single Family" ~ "Single Family", 
           type %in% c("Duplex", "Triplex", "Fourplex") ~ "Multiplex",
           grepl("unit", type, ignore.case = TRUE) ~ "Multiplex",
           TRUE ~ word(type, 1)), levels = type_order),
         street_address = address,
         address = ifelse(is.na(unit), address, paste0(unit, "-", address)))

```
# Summary
This dataset contains **`r nrow(listings_all)` data points** on **`r listings$mls_no %>% unique() %>% length()` unique MLS listings**. The last update was performed on **`r as_date(max(listings$datetime, na.rm = TRUE))`**, and scraped **`r listings %>% filter(as_date(datetime) == as_date(max(datetime))) %>% pull(pid) %>% unique() %>% length()`** different listing pages.

# Volume
## Where are the Listings?
```{r Listing Location, fig.width = 8, fig.height = 6}

top_20_cities <- listings %>% 
  distinct(mls_no, .keep_all = TRUE) %>%
  filter(!is.na(city), 
         !is.na(type),
         status == "For Sale") %>%
  count(city) %>%
  arrange(-n) %>%
  head(20) %>%
  pull(city)

listings %>%
  filter(city %in% top_20_cities,
         !is.na(type),
         status %in% c("For Sale", "Sold")) %>%
  distinct(mls_no, status, .keep_all = TRUE) %>%
  mutate(postal_city = factor(city, levels = top_20_cities)) %>%
  ggplot(aes(x = postal_city)) +
  geom_bar(aes(fill = type)) +
  facet_wrap(~ status, ncol = 1) +
  scale_fill_discrete(name = "") +
  scale_x_discrete(name = "") +
  scale_y_continuous(name = "Listing Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## When are they Being Listed
```{r When are the listings, fig.width = 8, fig.height = 4}

listings %>%
  distinct(mls_no, .keep_all = TRUE) %>%
  filter(ymd(list_date) > ymd("2020-04-01"), 
         status == "For Sale") %>%
  ggplot(aes(x = ymd(list_date))) +
  geom_bar(aes(fill = type), position = "stack", width = 0.9) +
  scale_fill_discrete(name = "") +
  xlab("List Date") +
  scale_y_continuous(name = "Count")
```

## Listing By Status Per Day
```{r Status by Day, fig.width = 8, fig.height = 8}

listings %>%
  group_by(date = as_date(datetime), status) %>%
  summarize(count = n()) %>%
  filter(!is.na(status)) %>%
  ggplot(aes(x = date, y = count)) +
  geom_bar(stat = "identity", aes(fill = status), width = 0.9) +
  scale_fill_manual(values = status_colors, name = "", guide = FALSE) +
  facet_grid(status ~ .) +
  xlab("Date Scraped") +
  scale_y_continuous(name = "Count") +
  geom_text(aes(y = count + max(count) * 0.075 * sign(count), label = count), size = 2)
```

## Changes in Inventory
This figure shows the net change in inventory in a given day. It is calculated as new listings with a status of "For Sale" minus the sum of new listings with the statuses "Sold", "Expired", "Withdrawn", and "Cancelled". Listings with a status of "Pending" are ignored.
```{r Balance in Listings, fig.width = 10, fig.height = 4}
listing_type <- c("For Sale" = "enter", 
                  "Sold" = "exit", 
                  "Defunct" = "exit",
                  "Pending" = NA, 
                  "Withdrawn" = "exit", 
                  "Cancelled" = "exit", 
                  "Expired" = "exit")

volume_changes <- listings %>%
  mutate(status = as.character(status), 
         status_type = listing_type[status]) %>%
    filter(status != "Pending", !is.na(status_type)) %>%
  group_by(date = as_date(datetime), status_type) %>% 
  summarize(N = n()) %>%
  pivot_wider(id_cols = date, names_from = "status_type", values_from = "N", values_fill = list(N = 0)) %>%
  mutate(net_change = enter - exit)
  
listings %>%
  filter(status != "Pending") %>%
  group_by(date = as_date(datetime), status) %>%
  summarize(count = n()) %>%
  mutate(status = as.character(status),
    status_type = listing_type[status],
         count = case_when(
           status_type == "enter" ~ count,
           status_type == "exit" ~ -count
         )) %>%
  ggplot(aes(x = date, y = count)) +
  geom_col(aes(fill = status), na.rm = TRUE) +
  scale_fill_manual(values = status_colors, guide = FALSE) +
  geom_text(data = volume_changes, aes(y = enter + max(enter) * 0.075, label = enter), size = 3) +
  geom_text(data = volume_changes, aes(y = -(exit + max(exit) * 0.075), label = exit), size = 3) +
  geom_line(data = volume_changes, aes(y = zoo::rollmean(net_change, 7, na.pad=TRUE, align = "right")), lty = 2, na.rm = TRUE) +
  labs(x = "Date", y = "Net Change in Active Listings")

```

# Pricing
## Per Square Foot in Different Regions
```{r Price per Area on vs. Off, fig.width=8, fig.height=8}

# Clean up the listings, bin each location into something meaningful
hrm_vs_other <- listings %>%
  filter(price < 1E6, between(sqft_mla, 450, 4000), # Nothing too fancy, thanks
         type %in% c("Single Family", "Condominium"))


# Fit each (location bin, type) to a linear model, extract the coefficients and make a pretty box for the graph
sqft_fits <- hrm_vs_other %>%
  group_by(loc_bin, type) %>%
  do(fit = lm(.$price ~ .$sqft_mla)) %>%
  mutate(slope = fit$coefficients[2],
         intercept = fit$coefficients[1],
         pretty = paste0("Base: ", scales::dollar(signif(intercept, 3)),
                         ifelse(is.na(slope), "", paste("\nPer Sq. ft.: ", scales::dollar(signif(slope, 3))))))

hrm_vs_other %>%
  mutate(days_since = as.numeric(ymd(list_date) - ymd("2020-04-01"))) %>%
  filter(days_since >= 0) %>%
  ggplot(aes(x = sqft_mla, y = price)) +
  geom_label(data = sqft_fits, aes(label = pretty),
             x = -Inf, y = Inf, hjust = 0, vjust = 1,
             size = 3) +
  geom_point(aes(color = days_since), na.rm = TRUE, alpha = 0.3) +
  labs(color = "", x = "Square Footage (MLS)", y = "List Price") +
  facet_grid(loc_bin ~ type) +
  scale_color_viridis_c(option = "plasma", 
                        direction = 1,
                        name = "Days Since\n2020-04-01") +
  scale_y_continuous(labels = scales::dollar) +
  geom_smooth(method = "lm", formula = y ~ x, na.rm = TRUE, color = "darkgrey")

```

## Price Trends by Listing Type
```{r Price per Square Foot Over Time, fig.width = 8, fig.height = 8}

sqft_vs_time <- listings %>%    
  filter(status %in% c("For Sale", "Sold"),
         between(sqft_mla, 450, 4000),
         !is.na(type))

svt_fit <- sqft_vs_time %>%
  group_by(status, type, loc_bin) %>%
  filter(n() > 20)


sqft_plot <- svt_fit  %>% 
  ggplot(aes(x = datetime, y = price/sqft_mla)) +
  geom_point(alpha = 0.1, aes(color = type)) +
  facet_grid(loc_bin ~ status, scales = "free_y") +
  geom_smooth(data = svt_fit, method = "loess", formula = y~x, aes(color = type), level = 0.01, lwd = 0.5) +
  labs(x = "List Date", y = "$/Sq. Ft.", color = "") +
  scale_color_brewer(name = "", palette = "Set1", direction = -1)

suppressWarnings(print(sqft_plot))

```


## Sale Price and Re-List Price
```{r price changes over time, fig.width = 8, fig.height = 4}
change_plot <- listings %>%
  group_by(pid) %>%
  filter(n() > 1) %>%
  arrange(datetime) %>%
  mutate(price_change = (price - lag(price))/lag(price)) %>%
  filter(!(is.na(price_change)),
         abs(price_change) < 0.5)

change_labels <- change_plot %>%
  group_by(status) %>%
  filter(datetime > now() - weeks(1),
         status %in% c("Sold", "For Sale")) %>%
  summarize(mean_change = mean(price_change))

change_plot %>%
  ggplot(aes(x = datetime, y = price_change)) +
  geom_jitter(data = filter(change_plot, price_change != 0), aes(color = status), width = 60*60*24*0.5, height = 0, alpha = 0.3) +
  geom_smooth(data = filter(change_plot, status %in% c("Sold", "For Sale")), 
              formula = y ~ x,
              method = "loess", 
              aes(color = status), 
              level = 0,
              lwd = 0.5) +
  scale_color_manual(values = status_colors, name = "") +
  labs(x = "Posting Date") +
  scale_y_continuous(name = "Price Change Since Last Event", labels = scales::percent, minor_breaks = seq(-1, 1, by = .05)) +
  geom_hline(yintercept = 0, lty = 2, lwd = 0.25) +
  scale_alpha_continuous(guide = FALSE) +
  #facet_grid(loc_bin ~ ., scales = "free_y") + 
  scale_x_datetime(limits = c(as_datetime(NA), now() + days(1))) +
  ggrepel::geom_label_repel(data = change_labels, x = now(), 
                            aes(y = mean_change, color = status, label = scales::percent(mean_change, accuracy = 0.01)), 
                            hjust = 0, direction = 'y')

```

## Assessment vs. Sale Price
```{r Assessment vs. Sale Price, fig.width = 8, fig.height = 5}

listings %>%
  filter(status == "Sold",
         assessment < 1E6, price < 2E6,
         assessment_year == "2020") %>%
  ggplot(aes(x = assessment, y = price)) +
  geom_point(aes(color = loc_bin), na.rm = TRUE) +
  scale_color_viridis_d(option = "plasma", name = "", end = 7/9) +
  scale_x_continuous(name = "Assessment (2020)", labels = scales::dollar) +
  scale_y_continuous(name = "Sale Price", labels = scales::dollar) +
  geom_abline(slope = c(1, 2, 3, 4, 5), 
              alpha = 1/c(1, 2, 3, 4, 5),
              lty = c(2, 3, 3, 3, 3))
```

# Listing-Level Analysis
## Changes in Price
These are the 20 most recently-listed addresses that have a change and have "Sold" somewhere in their recent history.
```{r Price Changes}
listings_with_change <- listings %>%
  filter(status != "Pending") %>%
  group_by(pid) %>%
  distinct(status, price, .keep_all = TRUE) %>%
  filter(length(unique(status)) > 1, # Has more than one datapoint
         (length(unique(price)) > 1 | status == "Sold")) %>% # Either it has a "Sold" data point or it has a price change
  arrange(list_date) %>%
  pull(pid)

close_order <- listings %>%
  filter(pid %in% listings_with_change) %>%
  group_by(pid, address) %>%
  summarize(list_date = max(datetime)) %>%
  arrange(list_date) %>%
  pull(address) %>%
  unique()

col_num <- 4
last_listings <- 20 
figure_rows <- last_listings %/% col_num + 1
```

```{r Price Changes Graph, fig.width = 10, fig.height = figure_rows + 2}

pid_to_address <- listings %>%
  group_by(pid) %>%
  summarize(address = head(address, 1))

pid_labeller <- pid_to_address$address
names(pid_labeller) <- as.character(pid_to_address$pid)

listings %>%
  filter(pid %in% tail(unique(listings_with_change), last_listings),
         price < 1E6) %>%
  arrange(list_date) %>%
  mutate(address = factor(address, levels = close_order)) %>%
  ggplot(aes(x = as_date(datetime), y = price)) +
  geom_line(aes(group = pid, color = status), lwd = 2, alpha = 0.5) +
  geom_point(aes(color = status), size = 4) +
  scale_color_manual(values = status_colors, name = "") +
  scale_x_date(name = "List Date") +
  facet_wrap(~ pid, scales = "free_y", ncol = col_num, labeller = labeller(pid = pid_labeller)) +
  ggrepel::geom_text_repel(aes(label = paste0("$", price/1000, "k"))) + 
  scale_y_continuous(name = "Price", breaks = seq(0, 1E6, by = 10000), labels = c()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Changes in Status
These are the 20 most recently-listed addresses that have a change. They don't necessarily have a "sold" entry in their history.
```{r Everything with Status Change, fig.width = 8, fig.height = 8}
listings_with_change <- listings %>%
  filter(status != "Pending") %>%
  group_by(pid) %>%
  distinct(status, price, .keep_all = TRUE) %>%
  filter(length(unique(status)) > 1) %>%
  pull(pid)

close_order <- listings %>%
  filter(pid %in% listings_with_change) %>%
  group_by(address) %>%
  summarize(list_date = max(datetime)) %>%
  arrange(list_date) %>%
  pull(address) %>%
  unique()

figure_rows <- last_listings %/% col_num + 1

```

```{r Status Change Figure, fig.width = 10, fig.height = figure_rows + 2}
listings %>%
  filter(pid %in% tail(unique(listings_with_change), last_listings),
         price < 1E6) %>%
  arrange(list_date) %>%
  mutate(address = factor(address, levels = close_order)) %>%
  ggplot(aes(x = as_date(datetime), y = price)) +
  geom_line(aes(group = pid, color = status), lwd = 2, alpha = 0.5) +
  geom_point(aes(color = status), size = 4) +
  scale_color_manual(values = status_colors, name = "") +
  scale_x_date(name = "List Date") +
  facet_wrap(~ pid, scales = "free_y", ncol = col_num, labeller = labeller(pid = pid_labeller)) +
  ggrepel::geom_text_repel(aes(label = paste0("$", price/1000, "k"))) + 
  scale_y_continuous(name = "Price", breaks = seq(0, 1E6, by = 10000), labels = c()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```

# Sale Price Model
```{r, Model setup, include=FALSE}
source("pricing-model.R")
rows_print <- 15
```

## Goodness of Fit
The current 'complex model' is **`r paste(deparse(complex_form, width.cutoff = 500), collapse="")`**.
```{r, model figs, fig.width = 12, fig.height = 5}
bind_cols(model = c("Dumb Model", "Complex Model"), 
          bind_rows(
  glance(dumb_model),
  glance(complex_model)
)) %>% as_tibble()

model_figs
```

## Model Terms
```{r model terms, rows.print = 20}
broom::tidy(complex_model) %>%
  mutate(estimate = scales::dollar(estimate),
         term = str_replace(term, "loc_bin", "") %>%
           str_replace(":style_bin", " // ") %>%
           str_replace(":sqft_dummy", " // sq.ft."),
         sig = symnum(p.value, corr = FALSE, na = FALSE, cutpoints = c(0, 
    0.001, 0.01, 0.05, 0.1, 1), symbols = c("***", "**", "*", ".", " "))) %>%
  select(term, estimate, sig, p.value)
```

## Undervalued by Model
Top `r rows_print` properties that are either undervalued by the model, or terrible deals.
```{r overvalued properties, rows.print = 10}

print_listing_table <- function(df) {
  df %>%
    mutate(scrape_date = as_date(datetime),
      street = ifelse(is.na(unit), street, paste0(unit, "-", street)),
           address_link = cell_spec(street,"html", link = url),
           value_score = round(xsv_z * 100),
           complex_prd = scales::dollar(complex_prd),
            price =  scales::dollar(price),
           loc_bin = str_trunc(case_when(
             postal_first %in% names(peninsula_codes) ~ peninsula_codes[postal_first],
             TRUE ~ city
           ), 16)) %>%
    select(`Date` = scrape_date,
           Address = address_link,
           Location = loc_bin,
           `Pred. Sale Price` = complex_prd,
           `Listing Price` = price,
           `Sq. Ft` = sqft_tla,
           `Value Score` = value_score) %>%
    kable("html", escape = FALSE) %>%
    kable_styling(bootstrap_options = c("hover", "condensed"))
  
}

test_set %>%
  arrange(xsv_z) %>%
  filter(list_date > (today() - days(14))) %>%
  head(rows_print) %>%
  print_listing_table(.)
```

## Overvalued by Model
Top `r rows_print` properties that are either overvalued by the model, or great deals!
```{r undervalued properties, rows.print = 10}

test_set %>%
  arrange(-xsv_z) %>%
  filter(list_date > (today() - days(14))) %>%
  head(rows_print) %>%
  print_listing_table(.)
```

## Listings on Gladstone
All of the listings on Gladstone St in Halifax
```{r Listings at on Gladstone}
test_set %>%
  arrange(desc(datetime)) %>%
  filter(grepl("Gladstone", address)) %>%
  print_listing_table(.)

```

## Listings on the Peninsula
The last 15 listings on the Halifax Peninsula by scrape date
```{r Listings on the Peninsula}
test_set %>%
  arrange(desc(datetime)) %>%
  filter(loc_bin == "Halifax Peninsula" | loc_bin %in% peninsula_codes) %>%
  head(15) %>%
  print_listing_table(.)

```